Efficient simulation of individual-based population models

### 2024-03-29 article.qmd
rm nonumber, write description

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université, CNRS, Institut du Risque et de l'Assurance 
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
date: last-modified
date-modified: last-modified
description: |
    TODO
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: default
  computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

The population at a given time $t$ is a random set
$$
    Z_t=\{ I_k \in \mathcal{I}; \; k= 1,\dots, N_t\},
$$
composed of all individuals (alive or dead) who have lived in the population before time $t$. As a random set, $Z_t$ can be represented by a random counting measure on $\mathcal{I}$ , that is an integer-valued measure $Z: \Omega \times \mathcal{I}\to \bar {\mathbb{N}}$ where for $A \in \mathcal{I}$, $Z(A)$ is the (random) number of individuals $I$ in the subset $A$. With this representation: 
$$
\begin{aligned}
    Z_t (\mathrm{d}\tau^b, \mathrm{d}\tau^d , \mathrm{d}x) &= \sum_{k=1}^{N_t} \delta_{I_k} (\tau^b, \tau^d,x), \nonumber \\
    \text{ with }  \int_{\mathcal{I}} f(\tau^b, \tau^d, x) Z_t (\mathrm{d}\tau^b, \mathrm{d}\tau^d , \mathrm{d}x) &= \sum_{k=1}^{N_t} f(I_k).
\end{aligned}
$$ {#eq-popZ}

@@

- When the event intensity does not depend on the population state, 
$$
    \big(\Lambda^e_t (Z_t)\big)_{t\in [0,T]} = \big(\mu^e(t)\big)_{t \in [0,T]},
$$ {#eq-poisson-intensity}
with $\mu^e$ a deterministic function, the events of type $e$ occur at the jump times of an inhomogeneous Poisson process of intensity function $(\mu^e(t))_{t \in [0,T]}$. When such an event occurs, the individual to whom the event happens to is drawn uniformly among alive individuals in the population.
In a given model, the set of events $e\in E$ with Poisson intensities will be denoted by $\mathcal{P}$.
- Otherwise, we assume that the global intensity $\Lambda^e_t(Z_t)$ at which the events of type $e$ occur in the population can be written as the sum of individual intensities $\lambda^e_t(I,Z_t)$:
$$
\begin{aligned}
    &\Lambda^e_t (Z_t) = \sum_{k=1}^{N_t} \lambda^e_t ( I_k,Z_t),  \\
    & \nonumber \text{with } \mathbb P\big( \text{event } e \text { occurs to an individual } I \in (t,t+\mathrm{d}t] | \mathcal{F}_t \big) \simeq  \lambda^e_t (I,Z_t)\mathrm{d}t.
\end{aligned}
$$ {#eq-global-intensity}

@@

**Step 2** The population size at the $n$--th event time $T_n$ is
$N_{T_n}$. The $(n+1)$--th event type and the individual to which this
event occur are thus chosen in the set
$$\mathcal{J}_n := \{ 1,\dots, N_{T_n}\} \times (\mathcal{E} \cup \mathcal{E}_W).$$
Conditionally to $\mathcal{F}_{T_n}$, let us first introduce the marked
Poisson measure $\bar Q^n$ on
$[T_n, \infty) \times \mathcal J_n \times \mathbb R^+$, of intensity:
$$
\begin{aligned}
\bar \mu^n(\mathrm{d}t, \mathrm{d}k, \mathrm{d}e , \mathrm{d}\theta ) & := \mathbf{1}_{\{t > T_n \}}\bar \Lambda (N_{T_n})\mathrm{d}t  \frac{\bar \lambda^e }{\bar \Lambda(N_{T_n})} \delta_{\mathcal J_n}(\mathrm{d}k, \mathrm{d}e) \frac{1}{\bar \lambda^e} \mathbf{1}_{[0,\bar \lambda^e]} (\theta)\mathrm{d}\theta,\\
\nonumber&  = \mathbf{1}_{\{t > T_n \}}\mathrm{d}t  \delta_{\mathcal J_n}(\mathrm{d}k, \mathrm{d}e) \mathbf{1}_{[0,\bar \lambda^e]}(\theta)\mathrm{d}\theta .
\end{aligned}
$$ {#eq-barmun}
By definition, $\bar Q^n$ has no jumps before $T_n$.\
As for the first event, the triplet $(T_{n+1}, E_{n+1}, I_{n+1})$ is
determined by the first jump of the measure
$Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e) := \int_{\mathbb R^+} \mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{T_n})\}}\bar Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta)$,
obtained by thinning of $\bar Q^n$. Finally, since the population
composition is constant on $[T_n, T_{n+1}[$, $Z_t = Z_{T_n}$, the
population on $[0,T_{n+1}]$ is defined by: 
$$
\begin{aligned}
\nonumber Z_{t\wedge T_{n+1}}  &  = Z_{t \wedge T_n }  + \mathbf{1}_{\{t\geq T_{n+1}\}}\phi^{E_{n+1}}(T_{n+1}, I_{n+1}), \\
& = Z_{t \wedge T_n } + \int_{t \wedge T_n}^{t \wedge T_{n+1}} \int_{\mathcal{J}_n
\times \mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta\leq \lambda_s^e(I_k, Z_{s^-})\}}\bar Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ).
\end{aligned}
$$  {#eq-z-tn} 
Applying $n$ times @eq-z-tn yields
that: 
$$\begin{aligned}
Z_{t\wedge T_{n+1}} = Z_0  + \sum_{l=0}^n \int_{t \wedge T_l}^{t \wedge T_{l+1}}\int_{ \mathcal{J}_l \times  \mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta\leq \lambda_s^e(I_k, \tilde Z_{s^-})\}}\bar Q^l (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ).
\end{aligned}
$$

>>>>

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université, CNRS, Institut du Risque et de l'Assurance 
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
date: last-modified
date-modified: last-modified
description: |
    This document provides a full description of the Stochastic Individual-Based Models (IBMs) that can be implemented in the IBMPopSim package. A unified mathematical and simulation framework is given, with a detailed description of the simulation algorithm. Examples of applications for the package are also provided, showing the performance and flexibility of IBMPopSim. 
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: default
  computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

The population at a given time $t$ is a random set
$$
    Z_t=\{ I_k \in \mathcal{I}; \; k= 1,\dots, N_t\},
$$
composed of all individuals (alive or dead) who have lived in the population before time $t$. As a random set, $Z_t$ can be represented by a random counting measure on $\mathcal{I}$ , that is an integer-valued measure $Z: \Omega \times \mathcal{I}\to \bar {\mathbb{N}}$ where for $A \in \mathcal{I}$, $Z(A)$ is the (random) number of individuals $I$ in the subset $A$. With this representation: 
$$
\begin{aligned}
    Z_t (\mathrm{d}\tau^b, \mathrm{d}\tau^d , \mathrm{d}x) &= \sum_{k=1}^{N_t} \delta_{I_k} (\tau^b, \tau^d,x),  \\
    \text{ with }  \int_{\mathcal{I}} f(\tau^b, \tau^d, x) Z_t (\mathrm{d}\tau^b, \mathrm{d}\tau^d , \mathrm{d}x) &= \sum_{k=1}^{N_t} f(I_k).
\end{aligned}
$$ {#eq-popZ}

@@

- When the event intensity does not depend on the population state, 
$$
    \big(\Lambda^e_t (Z_t)\big)_{t\in [0,T]} = \big(\mu^e(t)\big)_{t \in [0,T]},
$$ {#eq-poisson-intensity}
with $\mu^e$ a deterministic function, the events of type $e$ occur at the jump times of an inhomogeneous Poisson process of intensity function $(\mu^e(t))_{t \in [0,T]}$. When such an event occurs, the individual to whom the event happens to is drawn uniformly among alive individuals in the population.
In a given model, the set of events $e\in E$ with Poisson intensities will be denoted by $\mathcal{P}$.
- Otherwise, we assume that the global intensity $\Lambda^e_t(Z_t)$ at which the events of type $e$ occur in the population can be written as the sum of individual intensities $\lambda^e_t(I,Z_t)$:
$$
\begin{aligned}
    &\Lambda^e_t (Z_t) = \sum_{k=1}^{N_t} \lambda^e_t ( I_k,Z_t),  \\
    &  \text{with } \mathbb P\big( \text{event } e \text { occurs to an individual } I \in (t,t+\mathrm{d}t] | \mathcal{F}_t \big) \simeq  \lambda^e_t (I,Z_t)\mathrm{d}t.
\end{aligned}
$$ {#eq-global-intensity}

@@

**Step 2** The population size at the $n$--th event time $T_n$ is
$N_{T_n}$. The $(n+1)$--th event type and the individual to which this
event occur are thus chosen in the set
$$\mathcal{J}_n := \{ 1,\dots, N_{T_n}\} \times (\mathcal{E} \cup \mathcal{E}_W).$$
Conditionally to $\mathcal{F}_{T_n}$, let us first introduce the marked
Poisson measure $\bar Q^n$ on
$[T_n, \infty) \times \mathcal J_n \times \mathbb R^+$, of intensity:
$$
\begin{aligned}
\bar \mu^n(\mathrm{d}t, \mathrm{d}k, \mathrm{d}e , \mathrm{d}\theta ) & := \mathbf{1}_{\{t > T_n \}}\bar \Lambda (N_{T_n})\mathrm{d}t  \frac{\bar \lambda^e }{\bar \Lambda(N_{T_n})} \delta_{\mathcal J_n}(\mathrm{d}k, \mathrm{d}e) \frac{1}{\bar \lambda^e} \mathbf{1}_{[0,\bar \lambda^e]} (\theta)\mathrm{d}\theta,\\
&  = \mathbf{1}_{\{t > T_n \}}\mathrm{d}t  \delta_{\mathcal J_n}(\mathrm{d}k, \mathrm{d}e) \mathbf{1}_{[0,\bar \lambda^e]}(\theta)\mathrm{d}\theta .
\end{aligned}
$$ {#eq-barmun}
By definition, $\bar Q^n$ has no jumps before $T_n$.\
As for the first event, the triplet $(T_{n+1}, E_{n+1}, I_{n+1})$ is
determined by the first jump of the measure
$Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e) := \int_{\mathbb R^+} \mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{T_n})\}}\bar Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta)$,
obtained by thinning of $\bar Q^n$. Finally, since the population
composition is constant on $[T_n, T_{n+1}[$, $Z_t = Z_{T_n}$, the
population on $[0,T_{n+1}]$ is defined by: 
$$
\begin{aligned}
 Z_{t\wedge T_{n+1}}  &  = Z_{t \wedge T_n }  + \mathbf{1}_{\{t\geq T_{n+1}\}}\phi^{E_{n+1}}(T_{n+1}, I_{n+1}), \\
& = Z_{t \wedge T_n } + \int_{t \wedge T_n}^{t \wedge T_{n+1}} \int_{\mathcal{J}_n
\times \mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta\leq \lambda_s^e(I_k, Z_{s^-})\}}\bar Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ).
\end{aligned}
$$  {#eq-z-tn} 
Applying $n$ times @eq-z-tn yields
that: 
$$\begin{aligned}
Z_{t\wedge T_{n+1}} = Z_0  + \sum_{l=0}^n \int_{t \wedge T_l}^{t \wedge T_{l+1}}\int_{ \mathcal{J}_l \times  \mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta\leq \lambda_s^e(I_k, \tilde Z_{s^-})\}}\bar Q^l (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ).
\end{aligned}
$$
